### 2013-03-25 doc/sosp13/abstract.tex

\Grappa serves both as a C++ user library and as a foundation upon which
higher level languages can be developed or adapted. \Grappa tolerates delays
to remote memory by multiplexing thousands of lightweight execution contexts
to each processor core; balances load via fine-grained distributed
work-stealing; takes full advantage of network characteristics by aggregating
smaller data requests into large ones; and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa
system, programming examples using the library interface, and compare
performance for several irregular benchmarks to hand-optimized MPI code and
the Cray XMT, a custom system used to target the real time graph analytics
market.

>>>>

\Grappa serves both as a C++ user library and as a foundation upon which
higher level languages can be developed or adapted. \Grappa tolerates delays
to remote memory by multiplexing thousands of lightweight workers
to each processor core; balances load via fine-grained distributed
work-stealing; takes full advantage of network characteristics by aggregating
smaller data requests into large ones; and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa
system,
%programming examples using the library interface,
and compare
performance for several irregular benchmarks to hand-optimized MPI code and
the Cray XMT, a custom system used to target the real time graph analytics
market.

### 2013-03-25 doc/sosp13/communication.tex

\Grappa's communication support has two layers: user-level messaging interface
based on active messages; and network-level transport that supports request
aggregation for better communication bandwidth.

% In order to mitigate the low message injection rate limits of commodity
% networks, Grappa's communication stack has two layers:

\paragraph{Active messages interface} At the upper (user-level) layer, \Grappa
implements asynchronous active messages~\cite{vonEicken92}. Each message
consists of a function pointer, an optional argument payload, and an optional
data payload.

% When a task sends a message, the message is linked to a send queue
% associated with the message's destination node and the task continues
% execution. Note that the message contents stay in the worker's stack until
% it is sent to the network interface buffer, to avoid multiple copies of the
% payload.

\paragraph{Message aggregation} In our experiments the vast majority of upper
layer message requests are smaller than 44 bytes. Our measurements confirm
manufacturers' published data [15]; with 44-byte packets, the available
bisection bandwidth is only a small fraction (3\%) of the peak bisection
bandwidth. As mentioned earlier, commodity networks including InfiniBand
achieves their peak bisection bandwidth only when the packet sizes are
relatively large --- on the order of multiple kilobytes. The reason for this
discrepancy is the combination of overheads associated with handling each
packet (in terms of bytes that form the actual packet, processing time at the
card and processing on the CPU within the driver stack). Consequently, to make
the best use of the network, we must convert small messages into large ones.

\paragraph{Message processing mechanics} Since communication is very frequent
in \Grappa, aggregating and sending messages efficiently is very important. To
achieve that, \Grappa makes careful use of caches, prefetching and lock-free
synchronization operations. \TODO{Does the explanation below need a figure?}

Outgoing messages are organized as an array of linked lists, with one linked
list per destination node in the system. Each core in a given node is
responsible for aggregating and sending the resulting message to a set of
destination nodes. The outgoing message lists are located in a region of
memory shared across all cores in a \Grappa node (thus enabling cores to peak
at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
node, and links the buffer into the corresponding linked list.

Each core has a system worker per destination node it is responsible for; this
system worker periodically checks whether messages to the corresponding
destination node is either large enough or has waited past a time-out period,
at which point the message is sent out. Aggregating and sending a message
involves manipulating a shared data-structure (the message list). This is done
using CAS (compare-and-swap) operations to avoid high synchronization costs.

% Threads that dispatch a message allocate a message buffer (typically on their
% stack) and dispatch a send request. \Grappa examines the destination system
% node for this field, and adds the message to a linked list of messages
% destined for that node. Periodically, or when this linked list grows large,
% the list is transferred using a CAS (compare-and-swap} operation to the
% processor on the originating system node that is responsible for managing all
% outbound communication requests to the specified destination node. This
% two-step process is used because it was determined empirically that we could
% only infrequently use processor coherent access requests because they remain a
% bottleneck on modern multicore hardware.

Each node has a region of memory with send buffers where the final aggregated
messages are built. These buffers are visible to the network card, and
messages can be sent with user-mode operations only. When the worker
responsible for outbound messages to a given system node has received a
sufficient number of message send requests, or a timeout is reached, the
linked list of messages is walked and messages are copied to a send buffer.
This process requires careful prefetching because most of the outbound
messages are \emph{not} in the processor cache at this time (recall that a
core can be aggregating messages originating from other cores in the same
node). Once the send buffer has been formed it is handed off to GASnet for
transfer to the remote system node. RDMA is used if the underlying network
supports it.

There are two useful consequences of forming the send buffer at the time of
message transmission instead of along the way, as individual upper layer
message send requests are received. First, as previously mentioned most of the
messages are not in the cache and prefetching is used to run-ahead in the
linked list of messages in order to avoid cache misses. But once the send
buffer is formed it is in the cache (for the most part). Hence, when it is
handed off to GASnet for transfer across the physical wire, the network card
can pull the message buffer from the processor cache instead of main memory,
which we have found speeds performance. The second consequence of this
decision is that we do not need to pre-allocate buffers for all destination
nodes in the system, as the buffer can be allocated on the fly. Nevertheless
we have found it efficient to build a flow-control like protocol of
outstanding message buffers between pairs of system nodes.

Once the remote system node has received the message buffer it is unpacked.
Upper level \Grappa messages are delivered to each core at the destination
system node by examining the message buffer at each processor core and then
handing the message buffer to the next core for delivery of messages destined
to that core.

>>>>

In order to mitigate the low message injection rate limits of commodity networks, Grappa's communication stack has two layers: one for user-level messages and one for network-level messages.

At the upper (user-level) layer, Grappa implements asynchronous active messages~\cite{vonEicken92}. Each message consists of a function pointer, an optional argument payload, and an optional data payload. When a task sends a message, the message is copied to a send queue associated with the message's destination and the task continues execution.  In our experiments the vast majority of upper layer message requests are smaller than 44 bytes. Our measurements confirm manufacturers' published data [15]; with 44-byte packets, the available bisection bandwidth is only a small fraction (3\%) of the peak bisection bandwidth.

To make the best use of the network, we must convert our small messages into large ones, and this is the responsibility of Grappa’s lower networking layer. As mentioned earlier, commodity networks including InfiniBand achieves their peak bisection bandwidth only when the packet sizes are relatively large --- on the order of multiple kilobytes. The reason for this discrepancy is the combination of overheads associated with handling each packet (in terms of bytes that form the actual packet, processing time at the card and processing on the CPU within the driver stack).

Grappa's lower network layer makes careful use of processor cache and coherent access behaviors.  Threads that dispatch a message allocate a message buffer (typically on their stack) and dispatch a send request.  Grappa examines the destination system node for this field, and adds the message to a linked list of messages destined for that node.  Periodically, or when this linked list grows large, the list is transfered using a coherent compare-and-swap operation to the processor on the originating system node that is responsible for managing all outbound communication requests to the specified destination node.  This two-step process is used because it was determined empirically that we could only infrequently use processor coherent access requests because they remain a bottleneck on modern multicore hardware.

Once the processor core responsible for outbound messages to a given system node has received a sufficient number of message send requests, or a timeout is reached, the linked list of messages is walked to form a send buffer.  This involves making a copy of the messages in memory.  Careful prefetching is used because most of the outbound messages are \emph{not} in the processor cache at this time.  Once the send buffer has been formed it is handed off to GASnet for transfer to the remote system node.  RDMA is used if the underlying network supports it.  There are two useful consequences of forming the message buffer at the time of message dispatch instead of along the way, as individual upper layer message send requests are received.  First, as previously mentioned most of the messages are not in the cache and prefetching is used to run-ahead in the linked list of messages in order to avoid cache misses.  But once the message buffer is formed it is in the cache (for the most part).  Hence, when it is handed off to GASnet for transfer across the physical wire, the network card can pull the message buffer from the processor cache instead of main memory, which we have found speeds performance.  The second consequence of this decision we do not need to pre-allocate buffers for all destination nodes in the system, as the buffer can be allocated on the fly.  Nevertheless we have found it efficient to build a flow-control like protocol of outstanding message buffers between pairs of system nodes.

Once the remote system node has received the message buffer it is unpacked.  Upper level Grappa messages are delivered to each processor at the system node by examining the message buffer at each processor core and then handing the message buffer to the next processor for delivery of messages destined to that processor.

### 2013-03-25 doc/sosp13/tasks.tex

\paragraph{Context switching} \Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes --- importantly, a context is saved in a single 64 cache line.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.

Since \Grappa keeps a very large number of active workers, their context data will not fit in cache. Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum. We determined empirically that prefetching the fourth worker in the scheduling order is sufficient. This makes context switching effectively free of cache misses.

>>>>

\paragraph{Context switching} \Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes.

%% No attempt is made to fit them within a single cache line unfortunately :-( Mark
%%% --- importantly, a context is saved in a single 64 byte cache line.

% Pretty sure this is not true anymore.  Even if it is, it's not important -Mark
%
%Since the compiler sees all calls to the context switch routine, we
%can save even less state. Our context switch routine appears to the
%compiler as inline assembly; we declare all the registers we need
%to save as ``clobbered'' by the inline assembly routine, and the
%compiler will issue its own save and restore code as needed. This allows the
%compiler to avoid saving any registers that are not used, or are used
%for temporary values that are not needed after the context switch.

Since \Grappa keeps a very large number of active workers, their context data will not fit in cache. Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum. We determined empirically that prefetching the fourth worker in the scheduling order is sufficient. Also determined empirically, we prefetch three cache lines from the stack for each worker.  This makes context switching effectively free of cache misses even to hundreds of thousands of threads.
